Variant calling with transformers

### 2022-03-15 manuscript/manuscript.tex

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics. Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants. Here we describe a new approach that replaces the handcrafted statistical methods with a single, end-to-end deep learning model. Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into predicted haplotype sequences. 
\end{abstract}

@@

Training tensors were generated by selecting regions 150bp in width from alignments. The encoding process included fetching all reads aligning to the desired region. If more reads were found then a specified maximum read depth (100 for all analyses included in this work), reads were randomly downsampled to not exceed the maximum size. Reads were sorted by the reference coordinate of the first aligned base. For each aligned base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No 'gap' tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.

@@

@@

 Convolutions - do they make sense?

@@

@@

 In his essay "The Bitter Lesson" Richard Sutton remarked that "methods that leverage general computation" often outperform methods with handcrafted statistical or heuristic techniques.  Our work continues in this tradition, replacing the myriad finely tuned algorithms with a single model, learned end-to-end.

>>>>

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics. Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants. Here we describe a new approach that replaces the handcrafted statistical methods with a single, end-to-end deep learning model. Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. Trained on whole genome sequences from Genome-in-a-Bottle, we demonstrate that this approach yields sensitivity and precision similar to other top-performing callers, and that 
\end{abstract}

@@

Training tensors were generated by selecting 150bp genomic regions in a biased manner described below. For each region, we fetch all overlapping reads and sort them by first aligned reference position. If more than 100 overlapping reads were found we randomly downsampled the set to not exceed the maximum size. For each base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No gap tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.

@@

 Things to look at:
 Difficult regions
 Complex variants
 phasing

@@

We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model. Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish the task. In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.

Our tools moves the deep learning part from just classification (real or false) of a previously detected variant to actual construction of the variants themselves.

Previous uses of deep learning for variant detection have employed Convolutional Neural Networks (CNNs). While CNNs have been used successfully in computer vision tasks, genomic data are fundamentally different. Specifically, nearby pixels in an image are much more likely to be part of the same structure than pixels that are more distant, and CNNs exploit this implicit structure by repeatedly applying convolutions to a small patch of pixels. In contrast, reads aligning to a genomic region do not share this structure. Two adjacent "pixels" (base calls) that align to the same genomic coordinate from different reads are just as likely to be from the same structure as any two randomly selected pixels that share the same alignment coordinate. In other words, permuting the order of reads in a genomic pileup does not impact the information content, but permuting rows of an image would likely degrade

@@

Since the introduction of the first Transformer model (Vaswani et al. 2017) many improvements have been proposed, both in the general case and for vision-specific tasks

Lots of transformer variants out there - swin transformers, fastformer, hierarchical blah blah... good to mention these as future directions.

@@

 In his essay "The Bitter Lesson" Richard Sutton remarked that "methods that leverage general computation" often outperform methods with handcrafted statistical or heuristic techniques.  Our work continues in this tradition, replacing the myriad finely tuned algorithms with a single, fully learned model.
